Species motif participation provides unique information about species risk of extinction

### 2023-05-09 manuscript/JAE_reply.tex

    \textbf{R:}

@@

    \textbf{R:}

@@

\section*{Reply to Reviewer \#1}

@@

    \textbf{R:}

@@

    \textbf{R:}

@@

                \textbf{R:}

@@

                \textbf{R:}
            The sentence the reviewer refer to is "How disturbances such as population decline or extinction lead to the loss of species via  direct and indirect effects (i.e., secondary extinctions) has long been a vibrant area of research (Curtsdotter et al., 2011a; Dunne & Williams, 2009; Eklöf \& Ebenman, 2006 Santos et al., 2021)." We can understand the reviewer´s confusion, but in fact species extinctions are both the disturbance and the response; an initial extinction of a species (i.e. the primary extinction) can, since species are connected to each other via for example consumer -- resource interactions, cause other species to extinct (i.e. secondary extinctions). We have now clarified this on line XX: 
            "The loss of a species in a food web can, due to the mutual dependencies among species, trigger a cascade of additional extinctions, so called secondary extinctions. A species' position within a food web is known to affect its risk of going secondarily extinct ~\citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}."

@@

             \textbf{R:}
             We have now clarified how we in this manuscript define a disturbance. This is done on line XX. " The loss of  basal level diversity typically causes declines in both abundance and richness of all types of consumers: herbivores, predators, parasitoids, etc. \citep{scherber2010bottom}. Given these far-reaching effects we here choose to analyse persistence after loss of basal resources, i.e., effects of bottom-up disturbances. We here refer to disturbance as loss of basal species. "

@@

                \textbf{R:}
            We agree with the reviewer that this was not so clearly expressed. We have now re-formulated this section and clarified with an example. This is now located at line XX and reads: "However, unlike in-degree and trophic level, which focus only on direct prey of the focal species or on the lengths of paths between the focal species and basal resources, motif participation also incorporates information on the focal species' interaction partners. Two species might both have three prey and a trophic level of 3, but they might have quite different motif participation if one species's prey are strict specialists and the others' prey are generalists. Importantly, since specialists are generally more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than a predator of generalists."

@@

                \textbf{R:}
            [[we could also just rephrase these as small-scale and large-scale properties. Then it's maybe more obvious why meso-scale is in the middle.]]
            We have now rephrased this part, now located on line XX: "Therefore, a species' motif participation profile captures the indirect interactions the species are affected by, thereby capturing important structures that are missed by global (network level) and local (species level) properties.  As such, motif participation may complement degree and trophic level to more fully predict species' extinction risk."

@@

                We have lightly rephrased the start of this paragraph (quoted below) and added some references to papers describing empirical networks with similar size and connectance to those we simulate.
                Also note that we compare our simlated results with results for a set of more recent, larger, and especially well-resolved empirical networks in the SI. Although some of the whole-network structural properties differed, the range of species motif participations in our simulated networks spanned the whole range of observed motif participation in these empirical networks.

@@

                \textbf{R:} Proportion of role is the proportion of the species' motif role made up by the particular motif in each panel. This misunderstanding seems linked to a general difficulty in understanding precisely how we define motif roles. To help remedy this, we have now added motif role vectors to Figure 1 to make it explicit how counts of motifs are translated into vectors of normalised motif frequencies. These frequencies are equivalent to the proportion of a species' total motif participation made up by each motif. We also now use the perminology ``proportion of motif in role'' and ``proportion of motif in network profile'' in the axis labels of Figs. 2, 3, and 5 for consistency.

@@

                \textbf{R:} Can the Reviewer provide a reference for their claim that apparent competition is destabilising? Is this a very general trend, or something that only happens in rare cases? Note that we do not claim that apparent competition is \emph{never} destabilising, only that more frequent participation in the apparent competition motif is associated with greater persistence on average. Moreover, we cannot think of a reason why a bottom-up disturbance (i.e., loss of some resources) would be destabilising in an apparent competition motif - surely it is more likely that a predator with two prey will persist after losing one than than two predator sharing one prey will persist if that prey is lost. We can see how it is possible for strong \emph{top-down} effects to destabilise an apparent competition motif (e.g., if a predator suddenly exerts much more pressure on one of the prey than the other), but this paper is concerned with bottom-up effects only. Given the earlier work showing that Bayesian networks capture the majority of secondary extinctions that occur in topological models that include both top-down and bottom-up effects, we can conclude that top-down effects are much less common as a cause of extinction than bottom-up effects. Of course they can still be equally interesting and may be strong in particular focal systems, but as our simulations do not reflect a specific community we have no reason to expect top-down effects to have a stronger role than in the networks used to validate the Bayesian network approach.

@@

                None of this is to discount the reality of top-down extinctions or their interest from a scientific standpoint. However, given that our interest is in bottom-up extinctions in particular and that we are up front about the absence of top-down extinctions in our model [[]]

                [[Add a comment about how strong top-down effects could change our interpretation to discussion?]]

@@

                \textbf{R:}
                [[refocus on motif participation is definitely a good idea. Though there were interaction effects with disturbance strength, so we can't leave that out.]]

>>>>

    \textbf{R:} We thank the Editor for the option to revise and resubmit and for the helpful feedback from all concerned. We have taken careful note of the Associate Editor's and Reviewers' comments, particularly with regard to clarity and our statistical methods, and have extensively revised the manuscript. As a result, we are confident that our aims and results are now clearer and that that introduction and discussion more clearly place our work within the context of the literature. We hope that the Editor will agree that our work now meets the high standards of the Journal of Animal Ecology.

    [[Some nice closing?]]

@@

    \textbf{R:} We thank the Associate Editor for his feedback and appreciate the points raised by the Reviewers. We have extensively revised our manuscript with regard to clarity and believe that the current version makes our main aim and conclusions clearer. We have also redone our analyses in line with the concern raised by Reviewer 1. These changes have, to our reading, greatly improved the manuscript and we are pleased to present this revision for new consideration.

@@

\section*{Reply to Reviewer \#1} [[Done?]]

@@

    \textbf{R:} We thank the Reviewer for taking the time to read the manuscript and for their thoughtful feedback. We have redone our analyses with binomial error distribution as suggested (though with glm rather than glmer as the random effects proved singular). Our main results did not change much with this improved approach, though some of our supplemental results (e.g., those relating to overall network structure) were simplified. We hope that the revised manuscript is now clearer and easier to read and that our statistical approach is more reasonable. Below, we respond to each of the Reviewer's comments (preceded by \textbf{R:}) in turn.

@@

    \textbf{R:} We thank the Reviewer for taking the time to read our manuscript and for the thoughtful feedback. We particularly appreciate the pointers where the structure of our manuscript was unclear an where it was difficult to extract our main aims and results. In the present revision we have substantially reworked the introduction and discussion and reordered the methods and results in order to present our work in a more streamlined and, hopefully, understandable manner. We believe that the manuscript has been greatly improved as a result. Note that, in response to a comment from the other Reviewer, we have also redone our analyses and some results have slightly changed from the previous submission. Below, we respond to each of the Reviewer's comments (preceded by \textbf{R:}) in turn.

@@

                \textbf{R:} We have thoroughly revised the summary to make it clearer that our focus is on species participation in motifs rather than the presence of motifs within a network. We hope that it now more-closely parallels the rest of the manuscript and is easier to follow.

@@

                \textbf{R:} The sentence the reviewer refers to is ``How disturbances such as population decline or extinction lead to the loss of species via  direct and indirect effects (i.e., secondary extinctions) has long been a vibrant area of research (Curtsdotter et al., 2011a; Dunne & Williams, 2009; Eklöf \& Ebenman, 2006 Santos et al., 2021).'' We can understand the reviewer´s confusion, but in fact species extinctions are both the disturbance and the response.
                An initial extinction of a species (i.e. the primary extinction) can, since species are connected to each other via  consumer--resource interactions, cause other species to go extinct (i.e. secondary extinctions). We have now clarified this on line XX:

                \begin{quotation}
                The loss of a species in a food web can, due to the mutual dependencies among species, trigger a cascade of additional extinctions, so called secondary extinctions. A species' risk of going secondarily extinct is known to depend on its position within a food web~\citep{Santos2021,curtsdotter2011robustness, dunne2009cascading, Eklof2006}.
                \end{quotation}

@@

             \textbf{R:}
             We have now clarified how we in this manuscript define a disturbance. This is done on line XX.

             \begin{quotation}
             Here, we focus on persistence after the loss of plants or other basal resources, i.e., the effects of bottom-up disturbances.
            \end{quotation}

@@

                \textbf{R:} We agree with the reviewer that this was not so clearly expressed. We have now re-formulated this section and clarified with an example, beginning at line XX.

                \begin{quotation}
                However, unlike in-degree and trophic level, which focus only on direct prey of the focal species or on the lengths of paths between the focal species and basal resources, motif participation also incorporates information on the focal species' interaction partners. Two species might both have three prey and a trophic level of 3, but they might have quite different motif participation if one species's prey are strict specialists and the others' prey are generalists. Importantly, since specialists are generally more vulnerable to extinction than generalists, the predator of specialists will be more likely to lose prey and go extinct itself than a predator of generalists.
                \end{quotation}

@@

                \textbf{R:} To avoid introducing extra terminology, we have replaced all instances of `global' and `local' with `large-scale' and `small-scale' respectively. We hope it is now easier to understand motifs as a meso-scale descriptor of network structure. See line XX, quoted below, for a short description. We have also, as described above, explained how motif participation captures information about a focal species' interaction partners, and thereby the potential for indirect interactions. We do not refer to second-step indirect interactions in order to avoid confusion with that term.

                \begin{quotation}
                Therefore, a species' motif participation profile captures the indirect interactions the species are affected by, thereby capturing important structures that are missed by global (network level) and local (species level) properties.  As such, motif participation may complement degree and trophic level to more fully predict species' extinction risk.
                \end{quotation}

@@

                We have lightly rephrased the start of this paragraph (quoted below) and added some references to papers describing empirical networks with similar size and connectance to those we simulate.
                Also note that we compare our simulated results with results for a set of more recent, larger, and especially well-resolved empirical networks in the SI. Although some of the whole-network structural properties differed, the range of species motif participations in our simulated networks spanned the whole range of observed motif participation in these empirical networks.

@@

                \textbf{R:} Proportion of role is the proportion of the species' motif role made up by the particular motif in each panel. This misunderstanding seems linked to a general difficulty in understanding precisely how we define motif roles. To help remedy this, we have now added motif role vectors to Figure 1 to make it explicit how counts of motifs are translated into vectors of normalised motif frequencies. These frequencies are equivalent to the proportion of a species' total motif participation made up by each motif. We also now use the perminology ``proportion of motif in role'' and ``proportion of motif in network profile'' in the axis labels of Figs. 2, 3, and 5 for consistency. We have also thoroughly revised the results section to emphasize our main findings and place supplemental results in the proper context.

@@

                \textbf{R:} Previous research has shown that omnivory can either be stabilising or destabilising, depending on the precise disturbance, strengths of interactions, etc. Our results, which deal specifically with persistence after bottom-up disturbance, add to that picture of complexity without contradicting earlier findings. Note that we do not say that omnivory is \emph{never} beneficial - only that in our networks, in response to our focal disturbance, is is generally detrimental. We also note that our results show apparent competition as associated with less persistence when disturbance is weak, which is compatible with the possibility that the Reviewer mentions above (though we do not specifically model prey switching, which seems to be what the Reviewer refers to).  In the revision we have expanded our discussion on how the strength and direction of motif participation-persistence relationships vary network-to-network (especially in networks with different connectance). We hope that this addition provides some of the nuance the Reviewer missed from the previous revision.

@@

                None of this is to discount the reality of top-down extinctions or their interest from a scientific standpoint. However, since the Bayesian network framework does not include top-down effects any comment we might make about how top-down effects might affect our results would necessarily be pure speculation. Instead, we have added a reference to another paper exploring the relationship between motif roles and persistence in a dynamical context. That paper, due to the computational limits inherent in the dynamical approach, uses only a single-species disturbance. It does, however, include top-down as well as bottom-up effects and may provide the information the Reviewer is seeking.

@@

                \textbf{R:} We have extensively revised the discussion to focus more clearly on the relationships between motif participation and persistence, although we note that some of these relationships depend on the strength of disturbance. As such, disturbance strength remains a strong secondary theme in the present version. We hope that this revision is easier to follow.
